Yeast RNA-seq 48 sample comparison

### 2018-01-18 datasets/RNA-Seq/yeast48-benchmarkFDR.Rmd
Add more diagnosticplots, more realistic fold change simulation

## Introduction

@@

## Set up workspace

@@

## Data download

@@

### Read data into R and create a count table

@@

## Analysis with DESeq2 & FDR Benchmarking

@@

### Full analysis (48 vs 48 non-null)

@@

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resFULL), pval=resFULL$pvalue, SE=resFULL$lfcSE, 
                      ind_covariate = resFULL$baseMean, effect_size=resFULL$log2FoldChange, 
                      test_statistic=resFULL$stat))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=1)

@@

@@

### Null Comparison (10 versus 10)

@@

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=0)

@@

@@

@@

### Non-null Comparison (10 versus 10 null with spike-in)

@@

# randomly sample a log2FC from truncated N(0,0.5) (greater than 0.5) 
log2FC <- rep(0, nrow(dds_null))
log2FC[DE] <- rtruncnorm(nDE, 0.5, sd=0.5) * sample(c(1,-1), size=nDE, replace=TRUE)
counts_new <- counts(dds_null)
counts_new[,colData(dds_null)$fake=="A"] <- counts(dds_null)[,colData(dds_null)$fake=="A"] * 2^log2FC
counts_new <- apply(counts_new, 2, as.integer)
counts(dds_null) <- counts_new

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=0)

@@

@@

ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() +
  xlim(0,0.2)
```

@@

### Null Comparison (5 versus 5)

@@

```{r}
# select a random subset of 10 WT samples
dds_null <- dds[,colData(dds)$condition == "Snf2"]
dds_null <- dds_null[,sample(1:ncol(dds_null), 10)]

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=1)

@@

### Non-null Comparison (5 versus 5 null with spike-in)

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth))

geneExp <- geneExp %>% na.omit() %>% dplyr::filter(ind_covariate>=1)

@@

ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)
```

@@

### Session information

>>>>

# Introduction

@@

# Set up workspace

@@

# Data download

@@

## Read data into R and create a count table

@@

# Analysis of full 48v48 with DESeq2 (non-null)

@@

## Set up DESeq2 object

@@

## Check assumptions

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resFULL), pval=resFULL$pvalue, SE=resFULL$lfcSE, 
                      ind_covariate = resFULL$baseMean, effect_size=resFULL$log2FoldChange, 
                      test_statistic=resFULL$stat,
                      pzero=rowSums(counts(dds)==0)/ncol(counts(dds))  ))

# filter NAs and those with less than 50% expressed 
geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

```
Check unimodal assumption (for ash).

```{r}
ggplot(data=geneExp, aes(test_statistic)) +
  geom_histogram(bins=30)
```

@@

## FDR benchmarking

@@

# Null Comparison (10 versus 10)

@@

## Run DESeq2 on subset

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      pzero=rowSums(counts(dds)==0)/ncol(counts(dds))  ))

# filter NAs and those with less than 50% expressed 
geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

@@

## Check assumptions

Check unimodal assumption (for ash).

```{r}
ggplot(data=geneExp, aes(test_statistic)) +
  geom_histogram(bins=30)
```

@@

## FDR Benchmark

@@

# Non-null Comparison (10 versus 10 null with spike-in)

Here we'll repeat the above, but adding in some DE genes so that we can evaluate the
true positive and false positive rates of each method.

## Run DESeq2 on subset
```{r}
nDE <- 500

@@

# randomly sample a log2FC from original FCs (without regard to DE)
log2FC <- rep(0, nrow(dds_null))
log2FC[DE] <- resFULL$log2FoldChange[DE]

counts_new <- counts(dds_null)
counts_new[,colData(dds_null)$fake=="A"] <- counts(dds_null)[,colData(dds_null)$fake=="A"] * 2^log2FC
counts_new <- apply(counts_new, 2, as.integer)
counts(dds_null) <- counts_new

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth,
                      pzero=rowSums(counts(dds)==0)/ncol(counts(dds))  ))

# filter NAs and those with less than 50% expressed 
geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

```
## Check assumptions

Check unimodal assumption (for ash).

```{r}
ggplot(data=geneExp, aes(test_statistic)) +
  geom_histogram(bins=30)
```

@@

## FDR benchmark

@@

# fpr vs tpr
ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() +
  xlim(0,0.2)

# alpha vs tpr
ggplot(data=roc,aes(alpha, TPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)

# alpha vs fpr
ggplot(data=roc,aes(alpha, FPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)+
  geom_abline(slope=1, linetype="dashed", color="grey")
```

@@

# Null Comparison (5 versus 5)

@@

## Run DESeq2 on subset
```{r}
# select a random subset of 10 WT samples
dds_null <- dds[,colData(dds)$condition == "Snf2"]
dds_null <- dds_null[,sample(1:ncol(dds_null), 10)]

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      pzero=rowSums(counts(dds)==0)/ncol(counts(dds))  ))

# filter NAs and those with less than 50% expressed 
geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)
```

## Check assumptions

Check unimodal assumption (for ash).

```{r}
ggplot(data=geneExp, aes(test_statistic)) +
  geom_histogram(bins=30)
```

@@

# Non-null Comparison (5 versus 5 null with spike-in)

Here we'll repeat the above, but adding in some DE genes so that we can evaluate the
true positive and false positive rates of each method.

## Run DESeq2 on subset

@@

```{r}
geneExp <- tbl_df(data.frame(geneName=rownames(resNULL), pval=resNULL$pvalue, SE=resNULL$lfcSE, 
                      ind_covariate = resNULL$baseMean, effect_size=resNULL$log2FoldChange, 
                      test_statistic=resNULL$stat,
                      qvalue=truth,
                      pzero=rowSums(counts(dds)==0)/ncol(counts(dds))  ))

# filter NAs and those with less than 50% expressed 
geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

@@

# fpr vs tpr
ggplot(data=roc,aes(FPR, TPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)

# alpha vs tpr
ggplot(data=roc,aes(alpha, TPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2)

# alpha vs fpr
ggplot(data=roc,aes(alpha, FPR, color=blabel)) +
  geom_point() +
  geom_line() + 
  xlim(0,0.2) +
  geom_abline(slope=1, linetype="dashed", color="grey")
```

@@

# Session information
